---
title: "hw05: Factor management and visualization"
author: "MielleM"
output:
  github_document:
    toc: true
    toc_depth: 4
always_allow_html: yes
---

## Getting started 

```{r getting started}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(maps))
library(extrafont)
```
  
## Part 1: Factor management 
#### Am I working with factors?

```{r test if continent is factor}
is.factor(gapminder$continent)
is.factor(gapminder$country)

glimpse(gapminder)
```
Yes, looks like both continent and country are factors. Using glimpse, I can confirm that those are the only two factors in the gapminder tibble which contains 1,704 observations of six variables. 

#### Drop factors and levels: Oceania

First, I'll remove the continent Oceania. It only contains two countries which makes it a less interesting comparison than the other continents. I'm going to skip some piping here even though it will make the code slightly longer, because I want to separate the data manipulation from the sanity checks.  
```{r remove Oceania entries}
# remove Oceania
drop_ocea <- gapminder %>% 
  filter(continent != "Oceania")

# check the number of rows
glimpse(drop_ocea)

# check levels of continent factor
unique(drop_ocea$continent)
```

from glimpse(), I can see that the tibble now only contains 1,680 observations, so we have successfully removed rows. From unique(), I can see that it is the Oceania entries that have been removed from the dataset but Oceania remains a level in the factor. 

```{r remove Oceania as a factor level}
# remove Oceania factor level
no_ocea <- drop_ocea %>%
  mutate(continent = fct_drop(continent))

# check if Oceania remains as a factor level
is.factor(no_ocea$continent)
unique(no_ocea$continent)
```

Great, looks like continent is still a factor and now Oceania has been removed as a level. For my check, I decided against using str() because it provides too much information when, as in this case, I have fairly targeted questions about the data. 


#### Reorder levels based on the data


As we can see from unique() above, the columns are ordered alphabetically, which is pretty arbitrary. I'm going to reorder the continents by standard deviation and pipe this into a violin plot.

```{r order violin plot by continent standard deviation of population}
no_ocea %>% 
  mutate(continent = fct_reorder(continent, pop, .fun = sd)) %>% 
  ggplot(aes(continent, pop, fill = continent)) +
  scale_y_log10() + 
  geom_violin() +
  labs(title = "Country-level population by continent, 1952 - 2007",
       subtitle = "Continents ordered from lowest to highest population standard deviation",
       x = "Continent",
       y = "Population")

```

We can also create a graph ordered by minimum population. 

```{r order violin plot by continent minimum population}
no_ocea %>% 
  mutate(continent = fct_reorder(continent, pop, .fun = min)) %>% 
  ggplot(aes(continent, pop, fill = continent)) +
  scale_y_log10() + 
  geom_violin() +
  labs(title = "Country-level population by continent, 1952 - 2007",
       subtitle = "Continents ordered from lowest to highest minimum population",
       x = "Continent",
       y = "Population")

```

This demonstrates something about the way ggplot2 assigns fill colors-- they're clearly linked to the order that variables will be plotted (e.g. the first entry will always be pink), not something inherent to the data. 


## Part 2: File I/O

#### Export data to .csv

Before I write the data frame to a csv, let's filter to get a more reasonable data set to work with. I'm going to filter to the Americas only, with data at 10 year intervals instead of 5. 

```{r filter americas data at 10 year intervals}

amer_7 <- gapminder %>% 
  filter(continent == "Americas") %>% 
  filter(str_detect(year, "7$"))
  
amer_7 %>% 
  ggplot(aes(country, lifeExp)) +
  geom_point() +
  coord_flip() + #flip axes
  labs(title = "Life expectancy in the Americas, 1957-2007",
       x = "Life expectancy",
       y = "Country") +
  theme_light()
```

When I plot the filtered data, it displays alphabetically -- not super helpful for understanding trends in the data. 

I'm going to order the country data by median life expectancy to get a better idea of overall trends. 

```{r order country by median life expectancy}
amer_7_ord <- amer_7 %>%   
  mutate(country = fct_reorder(country, lifeExp, .fun = median))  # reorder country by median lifeExp

amer_7_ord %>% 
  ggplot(aes(country, lifeExp)) +
  geom_point() +
  coord_flip() +
  labs(title = "Life expectancy in the Americas, 1957-2007",
       x = "Life expectancy",
       y = "Country") +
  theme_light()
```

This plot is way more helpful! 

Now, let's experiment with exporting the data frame I've created to a .csv. Will the ordering be preserved if I re-import and plot it? 

```{r export amer_7_ord to .csv}
write_csv(amer_7_ord, "amer_7_ord.csv") 
```

I'm going to re-import the same .csv and plot it in the same way to see if it retains the ordering. 

```{r re-import csv, test if order the same}
read_csv("amer_7_ord.csv") %>%  #import .csv
  ggplot(aes(country, lifeExp)) +
  geom_point() +
  coord_flip() +
  labs(title = "Life expectancy in the Americas, 1957-2007",
       x = "Life expectancy",
       y = "Country") +
  theme_light()
  

```

Nope, the ordering is not preserved by the .csv. 

#### Export data to RDS

Let's try using `saveRDS()` and `readRDS()`, and use identical() to test if the files are the same. If you want to run this segment of the code at home, you'll need to specify a different file directory for the output. 

```{r save/read RDS}
saveRDS(amer_7_ord, '/Users/miellemichaux/Documents/STAT54X/hw05/amer_7_ord.rds') 

amer_7_ordRDS <- readRDS('/Users/miellemichaux/Documents/STAT54X/hw05/amer_7_ord.rds') 

identical(amer_7_ord, amer_7_ordRDS) 

```
Yes, the files appear to be the same, but I'll plot the imported RDS just to be sure. 

```{r plot RDS import}
amer_7_ordRDS  %>% 
  ggplot(aes(country, lifeExp)) +
  geom_point() +
  coord_flip() +
  labs(title = "Life expectancy in the Americas, 1957-2007",
       x = "Life expectancy",
       y = "Country") +
  theme_light()
```

To summarize: RDS exports and imports preserve the factor order, but writing to a .csv does not. 


## Part 3: Visualization design
> Remake at least one figure or create a new one, in light of something you learned in the recent class meetings about visualization design and color. Maybe juxtapose your first attempt and what you obtained after some time spent working on it. Reflect on the differences. If using Gapminder, you can use the country or continent color scheme that ships with Gapminder. Consult the dimensions listed in All the Graph Things.

> Then, make a new graph by converting this visual (or another, if you'd like) to a plotly graph. What are some things that plotly makes possible, that are not possible with a regular ggplot2 graph?

#### Remake a figure using data viz principles

### Building a better scatterplot
First, my original plot. 

```{r original life expectancy in the americas plot}
amer_7_ord %>% 
  ggplot(aes(country, lifeExp)) +
  geom_point() +
  coord_flip() +
  labs(title = "Life expectancy in the Americas, 1957-2007",
       x = "Life expectancy",
       y = "Country") +
  theme_light()
```

Now, I'll remake the same plot in a cleaner and more engaging way. 

```{r remake life expectancy by country scatterplot}
amer_lifeExp_plot <- amer_7_ord %>% 
  ggplot(aes(lifeExp, country)) +
  geom_point(color = "darkcyan") +
  coord_flip() +
  labs(title = "Life expectancy in the Americas, 1957-2007",
       x = "Life expectancy",
       y = "Country") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, color = "grey50")) +
  theme(axis.text.y = element_text(color = "grey50")) +
  theme(panel.grid.major.x= element_blank(), 
        panel.grid.minor.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank()) +
  theme(text=element_text(size = 10, family = "Arial", color = "grey30"))

amer_lifeExp_plot 
  
```

In the new plot, I've removed most of the lines as they were unnecessary for understanding the data but provided visual clutter. I changed the text colors to provide a visual hierarchy that emphasizes the title and axis labels. I've added color to the points, but only to provide visual interest-- it could just as easily be made black and white. 


### Mapping in ggplot2 vs. plotly

I'm now introducing `plotly()`, which doesn't render in github markdown. Please see the [plotly md]() in my homework 5 repo for downloadable code chucks that will faithfully reproduce what you see here as plotly graphs. 

#### Data wrangling for mapping: 
Before I can map, I need some data! 

Quickly, I'm going to modify some ggplot map code that I wrote for the Gapminder dataset for a previous assignment. I really wanted to revisit a map similar to one I had already made for [hw04](https://github.com/STAT545-UBC-students/hw04-MielleM/blob/master/hw04.md) to see if I could improve its usability.

```{r set up Americas life expectancy data, geographic data}

world <- map_data("world")

americas07 <-gapminder %>%
  filter(year == 2007) %>%
  filter(continent == "Americas") %>% 
  rename(region = country) %>% 
  mutate(region = as.character(region)) %>% 
  mutate(region = ifelse(region == "United States", "USA", region))

americasgeog <- right_join(world, americas07, by = "region") 

```

#### Static map in ggplot2

```{r map of life expectancy in the americas, using ggplot2}
gg <- ggplot() + 
  geom_polygon(data = americasgeog,
               aes(x=long,
                   y = lat,
                   group = group,
                   fill = lifeExp,
                   text = paste(region, "</b> \n", round(lifeExp,1), "years"))) + 
  coord_map("mollweide") +
  scale_fill_distiller(palette = 4, direction = 1, "Life\nexpectancy") + 
  theme_void() +
  theme(panel.grid = element_blank()) + # remove x axis bar
  ggtitle("Life expectancy in the Americas, 2007")
gg
```

#### Interactive mapping with ggplotly

Here's the same map as above, but made interactive through the magic of ggplotly. Try hovering your mouse over a country to see the name and life expectancy for 2007. 

```{r interactive map of life expectancy in the Americas, using ggplotly}

ggplotly(gg, tooltip = "text") #hover labels life expectancy

```

I find the interactive potential of plotly to be quite exciting for its application to maps. In a static map, I would have an extremely messy time labeling all these countries. With an interactive map, the user can read the hover text if they want to, or zoom in to get a better look -- all without cluttering the map with a bunch of labels.




### Using plotly syntax
As a bit of a bonus exercise, I'm going to try using plotly syntax (instead of piping ggplot2 into ggplotly) to recreate the original scatter plot of life expectancy in the Americas. I won't spend as much time here making it look nice. 

```{r initial use of plotly to graph life expectancy in the Americas}
amer_7_ord %>% 
  plot_ly(x = ~country, 
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers") %>% 
    layout(xaxis = list())
```
Plotly adds an empty Afghanistan entry here-- perhaps because it's still an empty factor level in this dataset. This doesn't happen with ggplot2, so there must be something a bit different in the defaults for how plotly builds figures. 

Let's try using factor_drop() to remove empty factor levels, and unique() to make sure levels were removed.  

```{r remove empty factor levels from Americas data}
americas <- amer_7_ord %>% 
  mutate(country = fct_drop(country))

unique(americas$country)
```

Now that we know only countries from the Americas are levels in the country factor, I'll plot it again. 

```{r re-plot americas data without missing levels}
  americas %>% 
  plot_ly(x = ~country, 
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers") %>% 
    layout(xaxis = list())
```
This worked! Maybe in the future, I'll play around more with plotly syntax. 


## Part 4: Writing figures to file

> Use ggsave() to explicitly save a plot to file. Then use ![Alt text] (/path/to/img.png) to load and embed it in your report. You can play around with various options, such as:

  >  Arguments of ggsave(), such as width, height, resolution or text scaling.
  >  Various graphics devices, e.g. a vector vs. raster format.
  >   Explicit provision of the plot object p via ggsave(..., plot = p). Show a situation in which this actually matters.


To specify which plot will be saved, I'm going to add `plot =`. This is usually a good idea, as if I don't specify which plot, ggsave will use the most recent plot created. If I later re-order my code chunks or add another plot, it will now save the wrong one. In my opinion, it's better to explicitly name the desired plot, even if it's not totally necessary. 

If I don't specify a file destination, the exported files will end up in the same folder as my assignment 5 R project. I've specified my homework 5 repo as the destination for the png image, which I've linked to below. 

I don't think SVGs can be exported directly to github, so you'll have to take my word for it that the svg export completed successfully on my home computer. Notice that as SVGs are vector images, there's no need to include dimensions (width + height in units) or a resolution (e.g. 300 dots per inch). 

```{r using ggsave to save plots}

#png 
ggsave("lifeExp6x4.png", plot = amer_lifeExp_plot, device = "png", path = "https://github.com/STAT545-UBC-students/hw05-MielleM/blob/master/test_images/", width = 6, height = 4, units = "in", dpi = 300)

#svg
ggsave("lifeExp6x4.svg", plot = amer_lifeExp_plot, device = "svg")

#pdf
# ggsave("lifeExp6x4.pdf", plot = amer_lifeExp_plot, device = "pdf", dpi = 300)

```

Check out the [plot](https://github.com/STAT545-UBC-students/hw05-MielleM/blob/master/test_images/lifeExp6x4.png) that I uploaded to my github repo. 

The pdf export had some issues with recognizing font type, so I haven't uploaded it. 

## Thanks to: 

- [tidyverse cheat sheet](https://www.rstudio.com/resources/cheatsheets/)
- [forcats guide](https://www.rdocumentation.org/packages/forcats/versions/0.3.0/topics/fct_reorder)
- [filter help](https://sebastiansauer.github.io/dplyr_filter/)
- [flip graph axis](https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2)
- [RDS help](https://stackoverflow.com/questions/23721750/behavior-of-saverds-and-readrds-in-regard-to-objects-attributes)
- [ggplot2 label rotation](https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2)
- [plotly annotations](https://plot.ly/r/text-and-annotations/)
- [customization of ggplot2 themes](https://rpubs.com/tgwilson/ggplot-meets-lea-pica-and-domo)
- [customizing plotly maps](https://stackoverflow.com/questions/50007258/plotly-australia-choropleth-map?rq=1)
- [plotly cheat sheet](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf)
- [plotly help](https://www.rdocumentation.org/packages/plotly/versions/4.8.0)
- [ggplotly hover labels assistance](https://blog.cpsievert.me/2018/01/30/learning-improving-ggplotly-geom-sf/)
- [changing fonts](https://fibosworld.wordpress.com/2013/02/17/change-fonts-in-ggplot2-and-create-xkcd-style-graphs/)
- [ggplot2 colors](http://sape.inf.usi.ch/quick-reference/ggplot2/colour)
- [ggsave](https://ggplot2.tidyverse.org/reference/ggsave.html)