---
title: "hw05 plotly graphs"
author: "MielleM"
output:
  github_document:
    toc: true
    toc_depth: 4
always_allow_html: yes
---

## Getting started 

```{r getting started}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(maps))
```

### Interactive mapping

#### Map data setup 
```{r set up Americas life expectancy data, geographic data}

world <- map_data("world")

americas07 <-gapminder %>%
  filter(year == 2007) %>%
  filter(continent == "Americas") %>% 
  rename(region = country) %>% 
  mutate(region = as.character(region)) %>% 
  mutate(region = ifelse(region == "United States", "USA", region))

americasgeog <- right_join(world, americas07, by = "region") 

``` 

```{r map of life expectancy in the americas, using ggplot2}
gg <- ggplot() + 
  geom_polygon(data = americasgeog,
               aes(x=long,
                   y = lat,
                   group = group,
                   fill = lifeExp,
                   text = paste(region, "</b> \n", round(lifeExp,1), "years"))) + 
  coord_map("mollweide") +
  scale_fill_distiller(palette = 4, direction = 1, "Life\nexpectancy") + 
  theme_void() +
  theme(panel.grid = element_blank()) + # remove x axis bar
  ggtitle("Life expectancy in the Americas, 2007")
gg
```

#### Interactive map in with ggplotly

Here's the same map as above, but made interactive through the magic of ggplotly. Try hovering your mouse over a country to see the name and life expectancy for 2007. 

```{r interactive map of life expectancy in the Americas, using ggplotly}

ggplotly(gg, tooltip = "text") #hover labels life expectancy

```

I find the interactive potential of plotly to be quite exciting for its application to maps. In a static map, I would have an extremely messy time labeling all these countries. With an interactive map, the user can read the hover text if they want to, or zoom in to get a better look -- all without cluttering the map with a bunch of labels.




### Using plotly syntax
As a bit of a bonus exercise, I'm going to try using plotly syntax (instead of piping ggplot2 into ggplotly) to recreate the original scatter plot of life expectancy in the Americas. I won't spend as much time here making it look nice. 

```{r data wrangling for plotly life expectancy}
amer_7 <- gapminder %>% 
  filter(continent == "Americas") %>% 
  filter(str_detect(year, "7$"))

amer_7_ord <- amer_7 %>%   
  mutate(country = fct_reorder(country, lifeExp, .fun = median))
```

```{r initial use of plotly to graph life expectancy in the Americas}
amer_7_ord %>% 
  plot_ly(x = ~country, 
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers") %>% 
    layout(xaxis = list())
```
Plotly adds an empty Afghanistan entry here-- perhaps because it's still an empty factor level in this dataset. This doesn't happen with ggplot2, so there must be something a bit different in the defaults for how plotly builds figures. 

Let's try using factor_drop() to remove empty factor levels, and unique() to make sure levels were removed.  

```{r remove empty factor levels from Americas data}
americas <- amer_7_ord %>% 
  mutate(country = fct_drop(country))

unique(americas$country)
```

Now that we know only countries from the Americas are levels in the country factor, I'll plot it again. 

```{r re-plot americas data without missing levels}
  americas %>% 
  plot_ly(x = ~country, 
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers") %>% 
    layout(xaxis = list())
```
This worked! Maybe in the future, I'll play around more with plotly syntax. 


